---
title: "Advanced Real Data Examples: Zero-Inflation, Semicontinuous, and Longitudinal Growth"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Real Data Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates **SelectBoost.gamlss** on three additional *real* datasets:
- **Zero-inflated counts** (`pscl::bioChemists`): ZIP / ZINB
- **Semicontinuous** (`datasets::airquality` Ozone): ZAGA
- **Longitudinal growth with random effects** (`nlme::Orthodont`): random intercepts

```{r}
library(SelectBoost.gamlss)
library(gamlss)
```

> Optional engines: `install.packages(c("glmnet","grpreg","SGL","future.apply"))`

## 1. Zero-inflated counts — `pscl::bioChemists` (ZIP/ZINB)

```{r}
if (requireNamespace("pscl", quietly = TRUE)) {
  data("bioChemists", package = "pscl")
  bc <- pscl::bioChemists
  # Basic cleaning (keep complete cases for core vars)
  keep <- complete.cases(bc[, c("art","kid5","phd","fem","mar","ment","dept")])
  bc <- bc[keep, ]
  bc$fem <- factor(bc$fem); bc$mar <- factor(bc$mar); bc$kid5 <- factor(bc$kid5); bc$dept <- factor(bc$dept)

  # ZIP engine via stepwise (mu) + grouped/elastic features
  set.seed(10)
  fit_zip <- sb_gamlss(
    art ~ 1, data = bc, family = gamlss.dist::ZIP(),
    mu_scope = ~ kid5 + fem + mar + phd + ment + dept,
    engine = "grpreg", grpreg_penalty = "grLasso",
    B = 80, pi_thr = 0.6, pre_standardize = FALSE, parallel = "auto", trace = FALSE
  )
  head(selection_table(fit_zip))
} else {
  cat("Package 'pscl' not installed; skipping bioChemists example.\n")
}
```

### Side-by-side metrics (ZIP vs ZINB)
```{r}
if (requireNamespace("pscl", quietly = TRUE)) {
zip_g <- fit_zip$final_fit; zinb_g <- fit_zinb$final_fit
if (inherits(zip_g, 'gamlss') && inherits(zinb_g, 'gamlss')) {
  met <- data.frame(
    model = c('ZIP','ZINBI'),
    AIC = c(AIC(zip_g), AIC(zinb_g)),
    GAIC_k2 = c(gamlss::GAIC(zip_g, k = 2), gamlss::GAIC(zinb_g, k = 2)),
    deviance = c(zip_g$deviance, zinb_g$deviance)
  )
  print(met)
} else {
  cat('Final fits not found; metrics skipped.\n')
}} else {
  cat("Package 'pscl' not installed; skipping bioChemists example.\n")
}
```

### ZINB often handles overdispersion better than ZIP
```{r}
if (requireNamespace("pscl", quietly = TRUE)) {
  set.seed(11)
  fit_zinb <- sb_gamlss(
    art ~ 1, data = bc, family = gamlss.dist::ZINBI(),
    mu_scope = ~ kid5 + fem + mar + phd + ment + dept,
    engine = "grpreg", grpreg_penalty = "grLasso",
    B = 80, pi_thr = 0.6, pre_standardize = FALSE, parallel = "auto", trace = FALSE
  )
  head(selection_table(fit_zinb))
} else {
  cat("Package 'pscl' not installed; skipping bioChemists example.\n")
}
```

**Notes.** Many `art` values are zero; `ZINB` usually outperforms `ZIP` when overdispersion is present. Group lasso treats factor levels as single grouped terms.

---

## 2. Semicontinuous (ZAGA) — `airquality::Ozone`

`Ozone` is non-negative and has zeros; positive part is continuous — ideal for **ZAGA** (zero-adjusted Gamma).

```{r}
aq <- subset(airquality, !is.na(Ozone) & !is.na(Wind) & !is.na(Temp) & !is.na(Solar.R))
aq$Month <- factor(aq$Month)

set.seed(20)
fit_zaga <- sb_gamlss(
  Ozone ~ 1, data = aq, family = gamlss.dist::ZAGA(),
  mu_scope    = ~ pb(Temp) + pb(Wind) + pb(Solar.R) + Month,
  sigma_scope = ~ pb(Temp),
  engine = "grpreg", grpreg_penalty = "grLasso",
  B = 80, pi_thr = 0.6, pre_standardize = TRUE, parallel = "auto", trace = FALSE
)
head(selection_table(fit_zaga))
```

### Effect plot for Temp on mu (ZAGA)
```{r}
if (requireNamespace('ggplot2', quietly = TRUE)) {
  print(effect_plot(fit_zaga, 'Temp', aq, what = 'mu'))
}
```

**Notes.** `ZAGA` models a point mass at zero plus a Gamma for the positive part. Group selection keeps whole spline bases together.

---

## 3. Longitudinal growth with random effects — `nlme::Orthodont`

Fit **random intercepts** by subject using `gamlss.random::random()`, while selecting fixed effects for μ.

```{r}
if (requireNamespace("nlme", quietly = TRUE) && requireNamespace("gamlss.random", quietly = TRUE)) {
  data(Orthodont, package = "nlme")       # distance ~ age, Sex, Subject
  od <- nlme::Orthodont
  od$Subject <- factor(od$Subject); od$Sex <- factor(od$Sex)

  set.seed(30)
  fit_long <- sb_gamlss(
    distance ~ gamlss.random::random(Subject),     # random intercept
    data = od, family = gamlss.dist::NO(),
    mu_scope = ~ pb(age) + Sex + pb(age):Sex,      # select fixed effects
    engine = "grpreg", grpreg_penalty = "grLasso",
    B = 80, pi_thr = 0.6, pre_standardize = TRUE, parallel = "auto", trace = FALSE
  )
  head(selection_table(fit_long))
```

### Effect plot for age on mu by Sex (create separate curves)
```{r}
if (requireNamespace('ggplot2', quietly = TRUE)) {
  gg <- effect_plot(fit_long, 'age', od, what = 'mu')
  print(gg)
}
} else {
  cat("Packages 'nlme' and/or 'gamlss.random' not installed; skipping longitudinal example.\n")
}
```

**Notes.** The `random(Subject)` term stays in the base formula (not in `mu_scope`) so it is always included, while SelectBoost focuses on selecting the **fixed** structure.
