---
title: "Comparing SelectBoost-GAMLSS Variants"
shorttitle: "Comparing SelectBoost-GAMLSS Variants"
author: 
- name: "Frédéric Bertrand"
  affiliation: 
  - Cedric, Cnam, Paris
  email: frederic.bertrand@lecnam.net
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing SelectBoost-GAMLSS Variants}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
#file.edit(normalizePath("~/.Renviron"))
LOCAL <- identical(Sys.getenv("LOCAL"), "TRUE")
#LOCAL=TRUE
knitr::opts_chunk$set(purl = LOCAL)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette compares: plain `sb_gamlss`, the SelectBoost-integrated `SelectBoost_gamlss`,
grid-based `sb_gamlss_c0_grid` + `autoboost_gamlss`, and the lightweight `fastboost_gamlss`.

```{r, cache=TRUE, eval=LOCAL}
library(gamlss)
library(SelectBoost.gamlss)

set.seed(1)
n <- 500
x1 <- rnorm(n); x2 <- rnorm(n); x3 <- rnorm(n); x4 <- rnorm(n)
mu <- 0.5 + 1.2*x1 - 0.8*x3
y  <- gamlss.dist::rNO(n, mu = mu, sigma = 1)
dat <- data.frame(y, x1, x2, x3, x4)

# Baseline
b0 <- sb_gamlss(
  y ~ 1, data = dat, family = gamlss.dist::NO(),
  mu_scope = ~ x1 + x2 + x3 + x4, sigma_scope = ~ x1 + x2,
  B = 50, pi_thr = 0.6, pre_standardize = TRUE, trace = FALSE
)
plot_sb_gamlss(b0)

# SelectBoost integration (single c0)
sb <- SelectBoost_gamlss(
  y ~ 1, data = dat, family = gamlss.dist::NO(),
  mu_scope = ~ x1 + x2 + x3 + x4, sigma_scope = ~ x1 + x2,
  B = 50, pi_thr = 0.6, c0 = 0.5, pre_standardize = TRUE, trace = FALSE
)
summary(sb) |> plot()

# c0 grid + autoboost
g <- sb_gamlss_c0_grid(
  y ~ 1, data = dat, family = gamlss.dist::NO(),
  mu_scope = ~ x1 + x2 + x3 + x4, sigma_scope = ~ x1 + x2,
  c0_grid = seq(0.2, 0.8, by = 0.2), B = 40, pi_thr = 0.6, pre_standardize = TRUE, trace = FALSE
)
plot(g)
confidence_table(g) |> head()

ab <- autoboost_gamlss(
  y ~ 1, data = dat, family = gamlss.dist::NO(),
  mu_scope = ~ x1 + x2 + x3 + x4, sigma_scope = ~ x1 + x2,
  c0_grid = seq(0.2, 0.8, by = 0.2), B = 40, pi_thr = 0.6, pre_standardize = TRUE, trace = FALSE
)
attr(ab, "chosen_c0")
plot_sb_gamlss(ab)

# fastboost (lightweight)
fb <- fastboost_gamlss(
  y ~ 1, data = dat, family = gamlss.dist::NO(),
  mu_scope = ~ x1 + x2 + x3 + x4, sigma_scope = ~ x1 + x2,
  B = 30, sample_fraction = 0.6, pi_thr = 0.6, pre_standardize = TRUE, trace = FALSE
)
plot_sb_gamlss(fb)
```



## Confidence Functionals (AUSC, Coverage, Quantiles, Weighted, Conservative)

We summarize the stability curves \(p_j(c0)\) into single-number scores:

- **AUSC**: normalized area under \(p_j(c0)\) across the `c0_grid`.
- **Thresholded AUSC** (\( (p - \pi^\star)_+ \) area): only counts confidence above the stability threshold.
- **Coverage**: fraction of `c0` where \(p \ge \pi^\star\).
- **Quantiles** (median/80th/90th): robust distributional summaries of stability.
- **Weighted AUSC**: apply a weight function \(w(c0)\) to emphasize parts of the grid.
- **Conservative AUSC**: use Wilson lower bounds for \(p\) before computing AUSC.

```{r, cache=TRUE, eval=LOCAL}
# Using the grid g created above
cf <- confidence_functionals(
  g, pi_thr = 0.6,
  q = c(0.5, 0.8, 0.9),
  weight_fun = function(c0) (1 - c0)^2,  # emphasize smaller c0
  conservative = TRUE, B = 40,           # use lower bounds
  method = "trapezoid"
)
head(cf)
plot(cf, top = 12, label_top = 8)

# Inspect stability curves of top terms
top_terms <- head(cf[order(cf$rank_score, decreasing = TRUE), c("term","parameter")], 4)
plot_stability_curves(g, terms = unique(top_terms$term), parameter = unique(top_terms$parameter)[1])
```


## Grouped penalties for factors & splines

```{r, cache=TRUE, eval=LOCAL}
library(gamlss)
library(SelectBoost.gamlss)

set.seed(42)
n <- 500
f  <- factor(sample(letters[1:3], n, TRUE))
x1 <- rnorm(n); x2 <- rnorm(n)
y  <- gamlss.dist::rNO(n, mu = 0.3 + 1.2*x1 - 0.7*x2 + 0.5*(f=="c"), sigma = exp(0.1 + 0.2*(f=="b")))

dat <- data.frame(y, f, x1, x2)

# Stepwise baseline
fit_step <- sb_gamlss(
  y ~ 1, data = dat, family = gamlss.dist::NO(),
  mu_scope = ~ f + x1 + pb(x2) + x1:f,
  sigma_scope = ~ f + x1 + pb(x2),
  B = 40, pi_thr = 0.6, pre_standardize = TRUE, trace = FALSE
)

# Group lasso for μ and sparse group lasso for σ
fit_grp <- sb_gamlss(
  y ~ 1, data = dat, family = gamlss.dist::NO(),
  mu_scope = ~ f + x1 + pb(x2) + x1:f,
  sigma_scope = ~ f + x1 + pb(x2),
  engine = "grpreg", engine_sigma = "sgl",
  B = 40, pi_thr = 0.6, pre_standardize = TRUE, trace = FALSE
)

plot_sb_gamlss(fit_step)
plot_sb_gamlss(fit_grp)
selection_table(fit_step) |> head()
selection_table(fit_grp)  |> head()
```


## Tuning & Group Knockoffs

```{r, cache=TRUE, eval=LOCAL}
# Tuning engines / penalties
cfgs <- list(
  list(engine="stepGAIC"),
  list(engine="glmnet", glmnet_alpha=1),
  list(engine="grpreg", grpreg_penalty="grLasso", engine_sigma="sgl", sgl_alpha=0.9)
)
base <- list(
  formula = y ~ 1, data = dat, family = gamlss.dist::NO(),
  mu_scope = ~ f + x1 + pb(x2) + x1:f, sigma_scope = ~ f + x1 + pb(x2),
  pi_thr = 0.6, pre_standardize = TRUE, sample_fraction = 0.7, parallel = "auto", trace = FALSE
)
tuned <- tune_sb_gamlss(cfgs, base_args = base, B_small = 30)
tuned$best_config

# Approximate group knockoffs for mu
sel_mu <- NULL
if (requireNamespace("knockoff", quietly = TRUE)) {
  sel_mu <- tryCatch(
    knockoff_filter_mu(dat, response = "y", mu_scope = ~ f + x1 + pb(x2), fdr = 0.1),
    error = function(e) {
      cat("Knockoff filter failed:", conditionMessage(e), "— skipping example.\n")
      NULL
    }
  )
} else {
  cat("Package 'knockoff' not installed; skipping knockoff example.\n")
}
if (!is.null(sel_mu)) sel_mu
```
