---
title: "Confidence Functionals for SelectBoost-GAMLSS"
shorttitle: "Confidence Functionals"
author: 
- name: "Frédéric Bertrand"
  affiliation: 
  - Cnam, Paris
  - Cedric
  email: frederic.bertrand@lecnam.net
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Confidence Functionals for SelectBoost-GAMLSS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
#file.edit(normalizePath("~/.Renviron"))
LOCAL <- identical(Sys.getenv("LOCAL"), "TRUE")
#LOCAL=TRUE
knitr::opts_chunk$set(purl = LOCAL)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

We consider stability curves \(p_j(c0)\) for each term \(j\) across a grid of SelectBoost thresholds `c0`. The following
functionals condense these curves into scalar confidences:

1. **AUSC** (Area Under the Stability Curve): normalized trapezoidal area over the grid.
2. **Thresholded AUSC**: integrate the positive excess \((p - \pi^\star)_+\).
3. **Coverage**: fraction of grid points with \(p \ge \pi^\star\).
4. **Quantiles**: median and high quantiles (e.g., 80th/90th) of \(p\) across `c0`.
5. **Weighted AUSC**: integrate \(w(c0)\,p(c0)\) normalized by \(\int w(c0)\,dc0\).
6. **Conservative AUSC**: replace \(p\) by a Wilson lower bound before integrating.

```{r}
library(gamlss)
library(SelectBoost.gamlss)

set.seed(1)
n <- 400
x1 <- rnorm(n); x2 <- rnorm(n); x3 <- rnorm(n)
y  <- gamlss.dist::rNO(n, mu = 1 + 1.3*x1 - 1.0*x3, sigma = 1)
dat <- data.frame(y, x1, x2, x3)

g <- sb_gamlss_c0_grid(
  y ~ 1, data = dat, family = gamlss.dist::NO(),
  mu_scope = ~ x1 + x2 + x3, sigma_scope = ~ x1 + x2,
  c0_grid = seq(0.2, 0.8, by = 0.2), B = 40, pi_thr = 0.6, pre_standardize = TRUE, trace = FALSE
)

cf <- confidence_functionals(
  g, pi_thr = 0.6, q = c(0.5, 0.8, 0.9),
  weight_fun = NULL, conservative = FALSE, method = "trapezoid"
)

plot(cf, top = 10, label_top = 6)
plot_stability_curves(g, terms = c("x1","x3"), parameter = "mu")
```
