---
title: "Real Data Examples with Different Distributions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Real Data Examples with Different Distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates **SelectBoost.gamlss** on real datasets using *different* distributions.
We include a **growth** example (Dutch boys) and further count/continuous/binary cases.

> You may want to install optional engines for best results:
> `install.packages(c("glmnet","grpreg","SGL","future.apply"))`

```{r}
library(SelectBoost.gamlss)
library(gamlss)
library(gamlss.data)
library(MASS)           # Insurance, quine
set.seed(1)
```

## 1. Growth data (BCT) — Dutch boys

We model **height** vs **age** with **BCT** (Box–Cox t). Grouped penalties select among smoothers and interactions with pubertal stage.

```{r}
data(boys, package = "gamlss.data")
boys <- subset(boys, !is.na(height) & !is.na(age) & !is.na(tanner))
boys$tanner <- factor(boys$tanner)

fit_growth <- sb_gamlss(
  height ~ 1, data = boys, family = gamlss.dist::BCT(),
  mu_scope    = ~ pb(age) + pb(age):tanner,
  sigma_scope = ~ pb(age),
  nu_scope    = ~ pb(age),
  engine = "grpreg", engine_sigma = "grpreg", engine_nu = "grpreg",
  grpreg_penalty = "grLasso",
  B = 80, pi_thr = 0.6, pre_standardize = TRUE,
  parallel = "auto", trace = FALSE
)

selection_table(fit_growth)[1:12, ]

# Effect plot for age on mu
if (requireNamespace('ggplot2', quietly = TRUE)) {
  print(effect_plot(fit_growth, 'age', boys, what = 'mu'))
}
# Stability curves for the most important terms (optional)
# g <- sb_gamlss_c0_grid(height ~ 1, data = boys, family = gamlss.dist::BCT(),
#                        mu_scope = ~ pb(age) + pb(age):tanner, sigma_scope = ~ pb(age),
#                        nu_scope = ~ pb(age), c0_grid = seq(0.2, 0.8, by = 0.1),
#                        B = 40, pi_thr = 0.6, pre_standardize = TRUE, parallel = "auto")
# plot_stability_curves(g, terms = c("pb(age)", "pb(age):tanner"), parameter = "mu")
```

**Interpretation.** Expect `pb(age)` to be stably selected for μ and σ; interactions like `pb(age):tanner` often appear around puberty.

---

## 2. Count data (PO) — Insurance claims

Fit **Poisson** with log-offset for exposure.

```{r}
data(Insurance, package = "MASS")
ins <- transform(Insurance, logH = log(Holders))

fit_po <- sb_gamlss(
  Claims ~ offset(logH), data = ins, family = gamlss.dist::PO(),
  mu_scope = ~ Age + District + Group,
  engine = "glmnet", glmnet_alpha = 1,    # lasso
  B = 100, pi_thr = 0.6, pre_standardize = FALSE,
  parallel = "auto", trace = FALSE
)

selection_table(fit_po)
```

---

## 3. Positive continuous (GA) — Old Faithful

Use **Gamma** for positive waiting times vs eruption length.

```{r}
data(faithful)
faith <- transform(faithful, eru2 = eruptions^2)

fit_ga <- sb_gamlss(
  waiting ~ 1, data = faith, family = gamlss.dist::GA(),
  mu_scope    = ~ pb(eruptions) + eru2,
  sigma_scope = ~ pb(eruptions),
  engine = "glmnet", glmnet_alpha = 0.5,   # elastic-net
  B = 60, pi_thr = 0.6, pre_standardize = TRUE,
  parallel = "auto", trace = FALSE
)

selection_table(fit_ga)

# Effect plot for eruptions on mu
if (requireNamespace('ggplot2', quietly = TRUE)) {
  print(effect_plot(fit_ga, 'eruptions', faith, what = 'mu'))
}
```

---

## 4. Binary (BI) — `mtcars` transmission

Treat `am` (0/1) as Bernoulli.

```{r}
mt <- transform(mtcars,
  am = as.integer(am),
  cyl = factor(cyl), gear = factor(gear), carb = factor(carb), vs = factor(vs))

fit_bin <- sb_gamlss(
  am ~ 1, data = mt, family = gamlss.dist::BI(),
  mu_scope = ~ wt + hp + qsec + cyl + gear + carb + vs,
  engine = "grpreg", grpreg_penalty = "grLasso",
  B = 80, pi_thr = 0.6, pre_standardize = TRUE,
  parallel = "auto", trace = FALSE
)

selection_table(fit_bin)
```

---

## 5. Overdispersed counts (NBII) — `quine` absences

**Negative Binomial II** handles extra-Poisson variation.

```{r}
data(quine, package = "MASS")
fit_nb2 <- sb_gamlss(
  Days ~ 1, data = quine, family = gamlss.dist::NBII(),
  mu_scope = ~ Eth + Sex + Lrn + Age,
  engine = "grpreg", grpreg_penalty = "grLasso",
  B = 80, pi_thr = 0.6, pre_standardize = FALSE,
  parallel = "auto", trace = FALSE
)

selection_table(fit_nb2)
```

---

### Tips
- If you have many factor levels or smoothers, prefer `engine="grpreg"`/`"sgl"` (group-aware).
- For very wide numeric designs, try `engine="glmnet"` with `glmnet_alpha` in \[0,1\].
- Use `pre_standardize=TRUE` (fast C++ scaler) when predictors are numeric and on different scales.
- For speed, enable parallel workers: `future::plan(multisession, workers = 4); parallel="auto"`.
